{\boldmath Pseudoscalar top-Higgs coupling: Exploration of $\mathrm{CP}$-odd observables to resolve the sign ambiguity }

### 2016-02-15 Paper_ttbarH_Draft.tex

%We also show that even when the number of events is reduced by an order of magnitude the most promising observables are still useful for testing the couplings
\end{abstract}
\maketitle
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Introduction}
\label{sec1}
After the discovery of a new boson $H$ by the ATLAS \cite{atlasH} and
CMS \cite{cmsH} collaborations, it has become of crucial importance
 to determine its physical
 properties with the highest possible precision.
 The study of the new boson's couplings
 to fermions is of great relevance
 and will allow us to better understand this particle's $\mathrm{CP}$-transformation
   properties, as well as the
extent to which this particle is consistent with the Higgs boson
predicted by the Standard Model (SM) of particle physics.
It is particularly important to test the coupling of the putative ({\bf (KK: at some point
  should we just start calling this the Higgs?  How do we make that transition?)}
Higgs boson to
the top quark.  This coupling governs
the main Higgs boson production mechanism (which proceeds via gluon fusion)
and it contributes to the
important Higgs boson decay mode to two photons.
It is also
involved in the scalar-field naturalness problem -- giving rise to the
leading dependence on the cut-off energy scale in the corrections to
the Higgs mass -- and it may play an important role in the EWSB {\bf KK: define EWSB?}
mechanism.  {\bf (KK: I reworded a lot of this paragraph -- is it still
  correct, or did I accidentally alter the meaning?)}\par

Given that the main Higgs boson production
process is dominated by a top quark loop and that
the diphoton and digluon decay channels are also mediated by a top
loop, these processes provide constraints on the scalar and
pseudoscalar $tH$ couplings, $\kp$ and $\kpt$
\cite{constraints1,constraints2, constraints3,constraints4}.
     {\bf (KK: OK to drop ``gluon fusion'' here, since we already said it was
       the main production process above?  Alternatively, we could say it here
       and skip it above.)}
     However,
these constraints assume that there are no other sources contributing
to the corresponding effective couplings; furthermore, in the case of the
diphoton decay channel (which also involves a $W$ boson loop), it is also
assumed that the
coupling of the Higgs boson to the $W$ is standard. In this sense, the
constraints derived from measurements of Higgs rates are indirect
constraints. Electric dipole moments (EDM) can also
impose stringent indirect constraints on $\kpt$ by assuming that there
are no new physics (NP) particles contributing to the loops of the
relevant diagrams and that the electron-Higgs coupling is that
predicted by the SM \cite{constraints1,edm}. In order to probe
the $tH$ coupling directly, processes with smaller cross sections need
to be taken into account.  \par

In contrast to the $\tau H$ coupling which can be studied trough the
decay $H\rightarrow \tau^+\tau^-$ \cite{tau}, the $tH$ coupling is
tested directly by making use of the production processes since the
phase space forbids the Higgs to decay to the pair $t\tbar$. The
production of the Higgs boson in association with the pair $t\tbar$
and the production in association with a single top or antitop are the
only two direct ways of probing the $tH$ coupling. The cross section
of the single top (antitop) production is smaller and it is given by
the interference between a diagram in which the Higgs is radiated from
the top (antitop) leg and one with the Higgs emmited from the
intermediate virtual $W$ boson. This then implies that the contraints
over $\kp$ and $\kpt$ derived from $tH$ and $\tbar H$ production are
dependent on the asummption made on the coupling of the Higgs to the
gauge boson $W$, $\kappa_W$. However, it is important to note that the
interference between the above mentioned diagrams present in $tH$
($\tbar H$) production can be exploited in order to determine the
relative sign between $\kp$ and $\kappa_W$ (see for example
\cite{tHmaltoni}). From the point of view of $t\tbar H$ production,
several observables sensitive to the couplings $\kp$ and $\kpt$ has
been proposed. Examples of such observables are the cross section, the
invariant mass distributions, the transverse Higgs momentum
distribution or the azymuthal-angle separation between $t$ and $\tbar$
just to name a few \citep{Guadagnoli}. An approach based on weighted
moments and optimal observables has been developed in \cite
{Gunion1,*Gunion2,*Gunion3,*Gunion} to discriminate the hypothesis of
a $\mathrm{CP}$-even Higgs from that of a $\mathrm{CP}$-mixed state
within the context of a $e^+ e^-$ as well as a $pp$ collider. All the
mentioned observables are $\mathrm{CP}$-even and hence are not
sensitive to relative sign between the scalar and pseudoscalar
couplings $\kp$ and $\kpt$. Such observables are quadratically
dependent on these couplings and then only provide an indirect measure
of $\mathrm{CP}$ violation. In order to be sensitive to the relative
sign between $\kp$ and $\kpt$, $\mathrm{CP}$-odd observables must be
considered. \par

Since the quark top decays before it can hadronize, its spin
information is passed on to the angular distributions of the decay
products in such a way that these particles work as spin analyzers
being the lepton the most powerful. On the other hand, it has been
shown that in $t\tbar$ production the top quark and antiquark spins
are strongly correlated \cite{Mahlon1,*Mahlon2,*Mahlon3,*atwood} and
this manifests in the double angular distributions of the decay
products of the $t$ and $\tbar$ systems. These correlations are
dependent in turn on the production mechanism and in the case of
$t\tbar H$ associated production are sensitive to the manner in which
the top couples to the Higgs boson. In fact, observables that exploit
the differences on the $t\tbar$ spin configurations are used in
\cite{Biswas} to improve the discrimination of the $t\tbar H$ signal
from the dominant irreducible background $t\tbar b\bbar$, in which the
Higgs boson is not involved. \par

In this paper, we define a set of observables linearly dependent on
$\kp$ and $\kpt$ and hence sensitive to the relative sign of these
couplings. The proposed observables are based on a particular set of
triple product structures (TP) that we extract naturally from the
expression of the differential cross section, which makes use in turn
of the fact that the kinematics of the decay products contain the
information on the $t$ and $\tbar$ spins and thus is sensitive to the
nature of the $tH$ coupling, just as was discussed in the previous
paragraph. By using spinor thechniques we simply relate the top quark
and antiquark spin vectors and the corresponding final state particle
momenta and separate the production process from the decay which
allows to identify straightforwardly the contributions linearly
sensitive to the couplings. Further, the TPs correlations in these
contributions incorporate the $t$ and $\tbar$ spin vectors. Starting
with these TPs, we not only recover the observables given in
\cite{Ellis,Guadagnoli} but also propose additional possibilities that
increase the sensitivity. In order to establish a hierarchy in the
sensitivity of the TPs under analysis we investigate three different
types of observables by using simulated events: asymmetries, mean
values and angular distributions. We note that TP correlations has
been used in \cite{Valencia1,*Valencia2} in the context of top-quark
production and decay and in \cite{Valencia3} in the framework of
anomalous color dipole operators.  \par

@@

At the LHC the $t\bar{t}H$ production proceeds via $q\bar{q}$
annihilation and $gg$ fusion processes. The relevant Feynman diagrams
at leading order are displayed in Fig.~\ref{fig1}, where the two first
rows correspond respectively to $q\bar{q}$ and $gg$ s-channels whereas
the last one depicts the $gg$ t-channels. By exchanging the gluon
lines in this row, three more $gg$-initiated diagrams can be
obtained. We describe the $tH$ cupling with the effective lagrangian

@@

expectation value, and the coefficients $\kp$ and $\kpt$ determine the
weigth of the scalar and pseudoscalar interaction, respectively. The

@@

and $\kpt\neq 0$ parameterizes a $\mathrm{CP}$-odd Higgs. \par

@@

\caption{Tree level Feynman diagrams contributing to $t\tbar H$ production at the LHC. Three more diagrams are obtained by exchanging the gluon lines in the t-channel diagrams.}

@@

Since the total cross-section is dominated by the $gg$ contribution,
let us focus on the $gg$-initiated production followed by the leptonic
SM decay of both $t$ and $\tbar$. By using the narrow-width

@@

cross section of the production of a top quark with spin vector $n_t$

@@

boson, whereas $d\Gamma(t(n_t)\to b{\ell}^+\nu_{\ell})$ and

>>>>

%We also show that even when the number of events is reduced by an order of magnitude the most promising observables are still useful for testing the couplings
\end{abstract}
\maketitle
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Introduction}
\label{sec1}
After the discovery of a new boson $H$ by the ATLAS \cite{atlasH} and
CMS \cite{cmsH} collaborations, it has become of crucial importance
 to determine its physical
 properties with the highest possible precision.
 The study of the new boson's couplings
 to fermions is of great relevance
 and will allow us to better understand this particle's $\mathrm{CP}$-transformation
   properties, as well as the
extent to which this particle is consistent with the Higgs boson
predicted by the Standard Model (SM) of particle physics.
It is of particular importance to test the coupling of the putative ({\bf (KK: at some point
  should we just start calling this the Higgs?  How do we make that transition?)}
Higgs boson to
the top quark.  This coupling governs
the main Higgs boson production mechanism (which proceeds via gluon fusion)
and it contributes to the
important Higgs boson decay mode to two photons.
It is also
involved in the scalar-field naturalness problem -- giving rise to the
leading dependence on the cut-off energy scale in the corrections to
the Higgs mass -- and it may play an important role in the EWSB {\bf KK: define EWSB?}
mechanism.  {\bf (KK: I reworded a lot of this paragraph -- is it still
  correct, or did I accidentally alter the meaning?)}\par

Given that the main Higgs boson production
process is dominated by a top quark loop and that
the diphoton and digluon decay channels are also mediated by a top
loop, these processes provide constraints on the scalar and
pseudoscalar $tH$ couplings, $\kp$ and $\kpt$
\cite{constraints1,constraints2, constraints3,constraints4}.
     {\bf (KK: OK to drop ``gluon fusion'' here, since we already said it was
       the main production process above?  Alternatively, we could say it here
       and skip it above.)}
     However,
these constraints assume that there are no other sources contributing
to the corresponding effective couplings; furthermore, in the case of the
diphoton decay channel (which also involves a $W$ boson loop), it is also
assumed that the
coupling of the Higgs boson to the $W$ is standard. In this sense, the
constraints derived from measurements of Higgs boson decay rates
{\bf (KK: correct to add ``boson decay'' here?)} are indirect
constraints. Electric dipole moments (EDMs) {\bf (KK: delete ``EDM'' since we
  never use it again?)} can also
impose stringent indirect constraints on $\kpt$ by assuming that there
are no new physics (NP) particles contributing to the loops of the
relevant diagrams and that the electron-Higgs coupling is that
predicted by the SM \cite{constraints1,edm}. In order to probe
the $tH$ coupling directly, processes with smaller cross sections need
to be taken into account.  \par

In contrast to the $\tau H$ coupling, which can be studied through the
decay $H\rightarrow \tau^+\tau^-$ \cite{tau},
the $tH$ coupling can only be tested directly via production processes, since the
Higgs boson is kinematically forbidden from decaying to a $t\tbar$ pair.
Two types of processes are of particular interest in this regard -- the
production of a Higgs boson in association with a $t\tbar$ pair
and in association with a single top or antitop. The cross section
for associated Higgs production with a single top (antitop)
{\bf (KK: did I phrase this correctly, calling it ``associated production'' in this way?)}
is smaller than that for production with a $t\tbar$ pair, and involves
the interference between a diagram in which the Higgs is radiated from
the top (antitop) leg and one with the Higgs emitted from the
intermediate virtual $W$ boson. Interestingly, this implies that the contraints
on $\kp$ and $\kpt$ derived from $tH$ and $\tbar H$ production are
dependent on the assumption made regarding the coupling of the Higgs boson to the
$W$ gauge boson, $\kappa_W$. Nevertheless, it is important to note that the
interference between the above mentioned diagrams
can be exploited to determine the
relative sign between $\kp$ and $\kappa_W$ (see for example
Ref.~\cite{tHmaltoni}).  Associated Higgs production with a
$t\tbar$ pair has been studied by several authors {\bf (KK: true?)}, and
various observables sensitive to the couplings $\kp$ and $\kpt$ have
been proposed. Examples of such observables (all of which are
$\mathrm{CP}$-even) are the cross section, 
invariant mass distributions, the transverse Higgs momentum
distribution and the azimuthal angular separation between the $t$ and $\tbar$,
to name a few \citep{Guadagnoli}. Also, an approach based on weighted
moments and optimal observables has been developed in Ref.~\cite
{Gunion1,*Gunion2,*Gunion3,*Gunion} to discriminate the hypothesis of
a $\mathrm{CP}$-even Higgs from that of a $\mathrm{CP}$-mixed state
within the context of an $e^+ e^-$ as well as a $pp$ collider.
{\bf (KK: is this last one CP even, too?)}  Now, $\mathrm{CP}$-even observables are not
sensitive to the relative sign between the scalar and pseudoscalar
couplings $\kp$ and $\kpt$. Such observables are quadratically
dependent on these couplings and thus only provide an indirect measure
of $\mathrm{CP}$ violation. In order to be sensitive to the relative
sign between $\kp$ and $\kpt$, $\mathrm{CP}$-odd observables must be
considered. \par

Since the top quark decays before it can hadronize, its spin
information is passed on to the angular distributions of its decay
products in such a way that these particles work as spin analyzers.
As is well known {\bf (ref)}, in the case of semileptonic top decay,
the charged lepton is the most powerful in this regard.
It is also known that the top quark and antiquark spins
are highly correlated in $t\tbar$ production\cite{Mahlon1,*Mahlon2,*Mahlon3,*atwood},
a feature that
is manifested in the double angular distributions of the decay
products of the $t$ and $\tbar$ systems {\bf (ref)}.
%The spin correlations
%between the $t$ and $\tbar$ are
%dependent in turn on the $t\tbar$ production mechanism, and in
In the case of
$t\tbar H$ associated production, the $t\tbar$ spin correlations
are also sensitive to the manner in which
the top couples to the Higgs boson. {\bf (KK: Did I simplify this too much?
  It seemed to repeat the previous idea a bit, but maybe I was misunderstanding it.)}
In fact, observables that exploit
the differences in the $t\tbar$ spin configurations were used in
Ref.~\cite{Biswas} to improve the discrimination of the $t\tbar H$ signal
from the dominant irreducible background $t\tbar b\bbar$, which does not
involve the Higgs boson. {\bf (KK: did I preserve the physics meaning
  correctly when I reworded this?)}

In this paper, we define a set of observables that are linearly dependent on
$\kp$ and $\kpt$ and are thus sensitive to the relative sign of these
couplings. The proposed observables are based on a particular set of
triple product (TP) structures that we extract naturally from the
expression for the differential cross section for {\bf (name of process)},
making use of the fact that the $t$ and $\tbar$ decay products
contain spin information and are sensitive to the
nature of the $tH$ coupling, as noted above.
By using spinor techniques we relate the top
and antitop spin vectors to final state particle
momenta and separate the production process from the decay which
allows to identify straightforwardly the contributions linearly
sensitive to the couplings. {\bf (KK: can we simplify the previous
  sentence?)}  Further, the TPs correlations in these
contributions incorporate the $t$ and $\tbar$ spin vectors. Starting
with these TPs, we not only recover the observables given in
\cite{Ellis,Guadagnoli} but also propose additional possibilities that
increase the sensitivity. In order to establish a hierarchy in the
sensitivity of the TPs under analysis we investigate three different
types of observables by using simulated events: asymmetries, mean
values and angular distributions. We note that TP correlations have
been used in \cite{Valencia1,*Valencia2} in the context of top-quark
production and decay and in \cite{Valencia3} in the framework of
anomalous color dipole operators.  \par

@@

At the LHC $t\bar{t}H$ production proceeds via $q\bar{q}$
annihilation and $gg$ fusion processes. The relevant leading-order Feynman
diagrams are displayed in Fig.~\ref{fig1}, with the first two
rows containing the $q\bar{q}$ and $gg$ $s$-channel diagrams, and
the last one depicting the $gg$ $t$-channel diagrams. By exchanging the gluon
lines in the third row, three more $gg$-initiated diagrams can be
obtained. We describe the $tH$ coupling with the effective Lagrangian

@@

expectation value, and the coefficients $\kp$ and $\kpt$ parameterize 
the scalar and pseudoscalar interaction, respectively. The

@@

and $\kpt\neq 0$ parameterize a $\mathrm{CP}$-odd Higgs. \par

@@

\caption{Tree-level Feynman diagrams contributing to $t\tbar H$ production
  at the LHC. Three more diagrams are obtained by exchanging the gluon
  lines in the $t$-channel diagrams.}

@@

In this section we focus on the gluon fusion contributions
to $t\tbar H$ production, since these dominate over the
the quark-antiquark annihilation contributions.  Furthermore,
we consider the case in which the top and antitop both decay
semileptonically.  By using the narrow-width

@@

cross section for the production of a top quark with spin vector $n_t$

@@

boson, and $d\Gamma(t(n_t)\to b{\ell}^+\nu_{\ell})$ and
